---
#title: "Assignment 3 ITI5145 - Qiroah App User Segmentation using Duolingo Growth Model"
#author: "Reza Darmawan"
#date: "Aug 2024"
output:
  html_document:
    toc: no
    df_print: paged
  pdf_document:
    number_sections: yes
    toc: no
header-includes:
- \usepackage{graphicx}
- \usepackage{color}
- \usepackage{multicol}
- \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, fig.pos = 'H')
```

# Assignment 3 ITI5145

### Qiroah App User Segmentation Using Duolingo Growth Model

## A. Preparation

### A.1. Libraries Used

```{r, warning=FALSE, results='hide'}

# create a function called install_and_load 
# as the name implied, it serves as a loader of required libraries
# and also installing them if it doesn't exist
install_and_load <- function(package) {
    
  # if the package is not loaded, 
  # hence the require function returns true  
  if (!require(package, character.only = TRUE)) {
    
    # install package in case it is not yet installed
    install.packages(package)
    
    # load the required library
    library(package, character.only = TRUE)
  }
}

required_packages<-c(
  'dplyr', # basic dataframe or tibble functionality
  'ggplot2', # to visualize the data (static)
  'readr', # read the file
  'tidyr', # package for data preprocessing
  'stringr', # library for dealing with strings
  'anytime', # to deal with unix epoch
  'DT', #datatable library
  'this.path' # for working directory setting
)

# Execute function to install and load packages 
# listed in required_packages vector
lapply(required_packages, install_and_load)
```

### A.2. Read Dataset

```{r}

# setting working directory to the path of this file----
cur_dir = dirname(this.path())
setwd(cur_dir)
getwd()

# Define string variable representing the dataset file name
file_dataset<-'Raw_Data_9_columns.csv'

text_dataset<-read_file(file.path(cur_dir,file_dataset))

#text_dataset<-iconv(text_dataset, from = "ISO-8859-1", to = "UTF-8")

# Use read_delim() function to read the dataset file
qiroah_app_events_data<-read_csv(text_dataset,
                                  
                        # the separator is comma (',') and the first row is the column name 
                        col_names=TRUE)

nrow(qiroah_app_events_data)
```

### A.3. Explore Dataset

```{r}
# Get a glimpse of events in the dataset
unique(qiroah_app_events_data$'Event Name')

# create a function to extract learning events

learning_events_extractor <- function(dataset){
  
  # Since we want to focus on the learning activity, the events will be filtered only based on 
  # certain categories: "lihat_video_preview", "rekam_suara_user","dengar_suara_user", "dengar_suara_ustadz_feedback", "ulangi_soal", "lihat_video_feedback","materi_selesai"
  
  learning_events<-c("lihat_video_preview", "rekam_suara_user","dengar_suara_user", 
                     "dengar_suara_ustadz_feedback", "ulangi_soal", "lihat_video_feedback","materi_selesai")
  
  # filter dataset to get only events related to learning
  learning_events_subset <- dataset %>% 
#  learning_events_subset <- dataset[, head(names(dataset), 3)] %>% 
    subset(dataset$'Event Name' %in% learning_events) %>%
    
    # convert number in Time column into date
    mutate(date_event = anydate(Time/1000)) %>%
    mutate(date_event = as.Date(paste(format(date_event, '%d'),
                              format(date_event, '%m'),
                              format(date_event, '%Y'), sep='/'), format='%d/%m/%Y'))
    
  # rename distinct ID column
  colnames(learning_events_subset)[colnames(learning_events_subset) == 'Distinct ID'] <- 'distinct_id'
  
  # return
  return(learning_events_subset)
  
}


qiroah_app_user_learning_events<-learning_events_extractor(qiroah_app_events_data)

# view filtered data
head(qiroah_app_user_learning_events) 
     #%>% arrange(desc(date_event)) %>% select(date_event))[[1]]


```

### A.4. User Extraction

```{r}
# get distinct user's distinct_id to get a list of Qiroah App users
qiroah_users<-data.frame(distinct_id=unique(qiroah_app_events_data$'Distinct ID'))

# see the top rows
head(qiroah_users)
```

## B. User Segmentation

### B.1 Create a function for user segmentation

```{r}

# create user segmentation function based on Duolingo growth model
user_segmentation <- function(user_vector, event_data, date_point_converted){
  
  # create summary of user event containing 2 recent activity date
  user_event_summary<-event_data %>%
    
    group_by(across(c('distinct_id','date_event'))) %>%
    
    summarise(a=n()) %>%
    
    group_by(distinct_id) %>%
    
    # get 2 recent events date
    summarise(most_recent_learning = max(date_event), 
             previous_most_recent_learning =nth(date_event, 2, 
                                                order_by = -as.numeric(date_event))) %>% 
    

    mutate(
      
      # add empty column for user segment
      user_segment='',
      
      # add new column for user segmentation date
      user_segmentation_date = date_point_converted)

  # iterate over user vector
  for (i in 1:nrow(user_event_summary)){
    
    if(!is.na(user_event_summary$previous_most_recent_learning[[i]])){
      
      # Current Users: learners active today, who were also active in the past week
      
      if(user_event_summary$most_recent_learning[[i]]==date_point_converted &
         user_event_summary$previous_most_recent_learning[[i]]>=date_point_converted-7){
        user_event_summary$user_segment[[i]]<-'Current Learning Users'
      } 
      
      # Reactivated Users: learners active today, who were also active in the past month (but not the past week)
      
      else if(user_event_summary$most_recent_learning[[i]]==date_point_converted &
         date_point_converted-30<=user_event_summary$previous_most_recent_learning[[i]] &
         user_event_summary$previous_most_recent_learning[[i]]<=date_point_converted-7){
        
        user_event_summary$user_segment[[i]]<-'Reactivated Learning Users'
        
      }
      
      # Resurrected Users: learners active today, who were last active >30 days ago
      
      else if(user_event_summary$most_recent_learning[[i]]==date_point_converted &
         user_event_summary$previous_most_recent_learning[[i]]<=date_point_converted-30){
        user_event_summary$user_segment[[i]]<-'Resurrected Learning Users'
      }
      
      # At Risk Weekly Active Users: learners who have been active within the past week, but not today
      
      else if(user_event_summary$most_recent_learning[[i]]!=date_point_converted &
         user_event_summary$most_recent_learning[[i]]>=date_point_converted-7){
        user_event_summary$user_segment[[i]]<-'At Risk Weekly Active Learning Users'
      }
      
      # At Risk Monthly Active Users: learners who were active within the past month, but not the past week
      
      else if(user_event_summary$most_recent_learning[[i]]>=date_point_converted-30 &
         user_event_summary$most_recent_learning[[i]]<=date_point_converted-7){
        user_event_summary$user_segment[[i]]<-'At Risk Monthly Active Learning Users'
      }
      
      # Dormant Users: learners who have been inactive for at least 30 days
      
      else if(user_event_summary$most_recent_learning[[i]]<=date_point_converted-30){
        user_event_summary$user_segment[[i]]<-'Dormant Learning Users'
      }
      
    }
    
    else {
      
      # New Users: learners who are experiencing Qiroah for the first time ever
      
      if(user_event_summary$most_recent_learning[[i]]==date_point_converted){
        user_event_summary$user_segment[[i]]<-'New Learning Users'
      }
      
      # At Risk Weekly Active Users: learners who have been active within the past week, but not today
      
      if(user_event_summary$most_recent_learning[[i]]!=date_point_converted &
         user_event_summary$most_recent_learning[[i]]>=date_point_converted-7){
        user_event_summary$user_segment[[i]]<-'At Risk Weekly Active Learning Users'
      }
      
      # At Risk Monthly Active Users: learners who were active within the past month, but not the past week
      
      else if(user_event_summary$most_recent_learning[[i]]>=date_point_converted-30 &
         user_event_summary$most_recent_learning[[i]]<=date_point_converted-7){
        user_event_summary$user_segment[[i]]<-'At Risk Monthly Active Learning Users'
      }
      
      # Dormant Users: learners who have been inactive for at least 30 days
      
      else if(user_event_summary$most_recent_learning[[i]]<=date_point_converted-30){
        user_event_summary$user_segment[[i]]<-'Dormant Learning Users'
      }
      
    }
    
  }
  
  return(user_event_summary)
  
}

# function to execute user segmentation on a certain date range
user_segmentation_range <- function(user_vector, event_data, date_point, day_backward=0){
  
  # convert date_point string to date
  date_point_converted<-as.Date(date_point, format='%Y-%m-%d')
  
  #filter event data only for events with date less than or equal to date point defined 
  event_data_filtered<-event_data %>% filter(date_event<=date_point_converted)
  
  # if the data is analyzed on user segmentation only for one day
  if(day_backward==0){
    
    # execute user segmentation for one day period only
    user_segmentation_result<-user_segmentation(user_vector,
                                                event_data_filtered, 
                                                date_point_converted) %>%
      
      # right join with user data
      right_join(user_vector,by=c('distinct_id')) %>%
      
      # add Non-Learning user category if users have no learning activity
      mutate_if(is.character,coalesce,'Non-Learning Users') %>%
      
      # add user segmentation date for users with no learning activity
      mutate_at(c('user_segmentation_date'), function(x) date_point_converted)
      #mutate_at(c('user_segmentation_date'), 
       #         as.Date(format(date_point_converted,'%d/%m/%Y'),'%d/%m/%Y'))
    
    # export user segmentation result to csv
    write.csv(user_segmentation_result,
              paste0("user_segment_",
                     format(date_point_converted, '%Y'),
                     format(date_point_converted, '%m'),
                     format(date_point_converted, '%d'),
                     ".csv"), row.names = FALSE)
    
  }
  
  # if the data is analyzed on user segmentation for a certain range
  else{
    
    # execute user segmentation for one day period only
    user_segmentation_result<-user_segmentation(user_vector,
                                                event_data_filtered, 
                                                date_point_converted) %>%
      
      # right join with user data
      right_join(user_vector,by=c('distinct_id')) %>%
      
      # add Non-Learning user category if users have no learning activity
      mutate_if(is.character,coalesce,'Non-Learning Users') %>%
      
      # add user segmentation date for users with no learning activity
      mutate_at(c('user_segmentation_date'), function(x) date_point_converted)
      #mutate_at(c('user_segmentation_date'), 
                #as.Date(format(date_point_converted,'%d/%m/%Y'),'%d/%m/%Y'))      
    
    # iterate user segmentation process on a given range of date
    for(i in 1:day_backward){
      
      # re-filter the event data based on the new date range
      event_data_refiltered<-event_data_filtered %>% filter(date_event<=date_point_converted-i)
      
      # execute user segmentation for one day period only
      user_segmentation_result_b<-user_segmentation(user_vector,
                                                    event_data_refiltered, 
                                                    date_point_converted-i) %>%
      
        # right join with user data
        right_join(user_vector,by=c('distinct_id')) %>%
      
        # add Non-Learning user category if users have no learning activity
        mutate_if(is.character,coalesce,'Non-Learning Users') %>%

        # add user segmentation date for users with no learning activity
        mutate_at(c('user_segmentation_date'), function(x) date_point_converted-i)
                # as.Date(format(date_point_converted-i,'%d/%m/%Y'),'%d/%m/%Y'))        
              
      # combine user segmentation dataframe
      user_segmentation_result<-rbind(user_segmentation_result, user_segmentation_result_b)
      
    }
    
    # export user segmentation result to csv
    write.csv(user_segmentation_result,
              paste0("user_segment_",
                     format(date_point_converted, '%Y'),
                     format(date_point_converted, '%m'),
                     format(date_point_converted, '%d'),
                     "_to_",
                     format(date_point_converted-day_backward, '%Y'),
                     format(date_point_converted-day_backward, '%m'),
                     format(date_point_converted-day_backward, '%d'),
                     ".csv"), row.names = FALSE)
    
  }
  
  return(user_segmentation_result)
  
}
```

### B.2. Execute User Segmentation (e.g. for 7 days prior since 1 August 2024)

```{r}

qiroah_user_segmentation<-user_segmentation_range(qiroah_users,qiroah_app_user_learning_events, '2024-08-01',7) 
```

### B.3. User Segmentation Visualization (e.g. per 1 August 2024)

```{r}

user_segmentation_capture<-function(segmentation_data,capture_date){
  
  segmentation_data %>%
  
    # filter user segmentation data 
    filter(user_segmentation_date==as.Date(capture_date,'%Y-%m-%d')) %>%
    
    # group by user segment
    group_by(user_segment) %>%
    
    # count number of user per segment
    summarise(count=n()) %>%
    
    # visualize in horizontal bar chart
    ggplot() +
    geom_col(aes(count, user_segment), fill = '#076fa2', width = 0.6) +
    labs(title = "Qiroah App User Segmentation from Learning Activity",
         subtitle=paste0("Cutoff Date: ",capture_date), 
         y='User Segments',x='Number of Users')
  
}

user_segmentation_capture(qiroah_user_segmentation,'2024-08-01')
```

```{r}
# Use read_delim() function to read the dataset file
qiroah_user_segmentation<-read_csv(file.path(cur_dir,'user_segment_20240801_to_20230802.csv'),
                                  
                        # the separator is comma (',') and the first row is the column name 
                        col_names=TRUE)

head(qiroah_user_segmentation)
```

### B.4 Create function to calculate user transition

```{r}
# this function will be used to calculate user retention
calculate_user_retention<-function(user_segmentation_data){
  
  # get earliest date in user segmentation data
  earliest_date<-min(user_segmentation_data$user_segmentation_date)
  
  # get latest date in user segmentation data
  latest_date<-max(user_segmentation_data$user_segmentation_date)
  
  # check if sufficient amount of data provided
  # in which the dataset should not only contain segmentation from one period
  if(earliest_date==latest_date) {
    
    print("Insufficient amount of observation provided: the dataset only contains user segmentation in one period.")
    
    # return empty dataframe
    return(data.frame(distinct_id=character(), 
                      user_retention_period=as.Date(character()), 
                      current_segmentation=character(), 
                      previous_segmentation=character()))
    
  
    } else { # proceed if dataset satisfies the criteria
      # convert date point defined in parameter as Date
      
      # filter the data for the current analysis period
      user_segment_data_current<-user_segmentation_data %>% filter(user_segmentation_date!=earliest_date) %>%
        
        # create a new column representing the current segmentation
        mutate(current_segmentation=user_segment,
               user_retention_period=user_segmentation_date)
      
      # filter the data for the previous analysis period
      user_segment_data_previous<-user_segmentation_data %>% filter(user_segmentation_date!=latest_date)%>%
        
        # create a new column representing the previous date segmentation
        mutate(previous_segmentation=user_segment,
               user_retention_period=user_segmentation_date+1)
      
      # join current user segmentation data with previous user segmentation data
      
      user_retention_data<-user_segment_data_current %>%
        inner_join(user_segment_data_previous, by=c('distinct_id','user_retention_period')) %>%
        
        # select relevant column
        select(distinct_id, user_retention_period, current_segmentation, previous_segmentation)
      
      user_retention_data$transition<-paste(user_retention_data$previous_segmentation,user_retention_data$current_segmentation,sep='--')
      
      # export user retention result to csv
      write.csv(user_retention_data,
                paste0("user_retention_",
                       format(latest_date, '%Y'),
                       format(latest_date, '%m'),
                       format(latest_date, '%d'),
                       "_to_",
                       format(earliest_date+1, '%Y'),
                       format(earliest_date+1, '%m'),
                       format(earliest_date+1, '%d'),
                       ".csv"), row.names = FALSE)
      
      return(user_retention_data)
      
    }
  }

```

### B.5. Execute User Transition Calculation (for 365 days prior since 1 August 2024)

```{r}
qiroah_user_retention<-calculate_user_retention(qiroah_user_segmentation)
```

```{r}
unique(qiroah_user_retention$transition)
```

### B.6. Create function for user segment summarization and visualization

```{r}

# function for user segment summarization
user_segment_summarization<-function(user_segment_data){
  
  # identify range of date
  earliest_date<-min(user_segment_data$user_segmentation_date)
  latest_date<-max(user_segment_data$user_segmentation_date)
  
  # group the segmentation data based on date and user segment
  user_segment_data_summary<-user_segment_data %>% group_by(across(c(
    user_segmentation_date,
    user_segment
    ))) %>%
    
    summarise(n_user=n())
  
  # export user retention result to csv
  write.csv(user_segment_data_summary,
            paste0("user_segmentation_summary_",
                   format(latest_date, '%Y'),
                   format(latest_date, '%m'),
                   format(latest_date, '%d'),
                   "_to_",
                   format(earliest_date+1, '%Y'),
                   format(earliest_date+1, '%m'),
                   format(earliest_date+1, '%d'),
                   ".csv"), row.names = FALSE)
  
  # return the data summary
  return(user_segment_data_summary)
}

# function to visualize the trend of a certain user segment 
user_segment_visualization<- function(user_segment_summary_data, selected_segment){
  
  # filter user segment data
  filtered_segment<-user_segment_summary_data %>% filter(user_segment==selected_segment)
  
  # identify range of data
  earliest_date<-format(min(filtered_segment$user_segmentation_date),'%Y-%m-%d')
  latest_date<-format(max(filtered_segment$user_segmentation_date),'%Y-%m-%d')
  
  # visualize a line chart
  filtered_segment %>%
    ggplot(aes(x=user_segmentation_date, y=n_user)) +
    geom_line(color="#69b3a2", size=1)+ 
    labs(title = "Qiroah App Learning Activity User Segment Trend",
         subtitle=paste0('User Segment: ',selected_segment,
                         ', Period: ',earliest_date,' to ',latest_date),
         y='Number of Users',x='Date')
}



```

### B.7. Execute User Segment Summarization and visualization

```{r}
# execute user segment summary
qiroah_user_segment_summary<-user_segment_summarization(qiroah_user_segmentation)

# execute visualization in iteration of user segment
for(i in unique(qiroah_user_segment_summary$user_segment)){
  print(user_segment_visualization(qiroah_user_segment_summary,i))
}
```

### B.8. Create a function for transition rate calculation and summarization

```{r}

# create a function to create a summarization of user retention
retention_rate_summarization<-function(user_segment_summary, user_retention_data){
  
  # identify range of date
  earliest_date<-min(user_retention_data$user_retention_period)
  latest_date<-max(user_retention_data$user_retention_period)
  
  # create a summary based on period and transition
  user_retention_summary<-user_retention_data %>% group_by(across(c(
    user_retention_period, previous_segmentation, current_segmentation
  ))) %>%
    summarise(n_user=n())
  
  head(user_retention_summary)

  # join user segment summary and user retention summary to get transition rate between segments
  user_retention_rate_data<-user_segment_summary %>% 
    
    # prepare join key column
    mutate(user_retention_period=user_segmentation_date+1,
           previous_segmentation=user_segment,
           previous_segment_count=n_user) %>%
    
    # select relevant column
    select(user_retention_period,
           previous_segmentation,
           previous_segment_count) %>%
    
    # right join to take all row from user retention summary
    right_join(user_retention_summary, by=c('user_retention_period','previous_segmentation')) %>%
    
    # select relevant columns
    select(user_retention_period, 
           previous_segmentation, 
           current_segmentation,
           n_user,previous_segment_count) %>%
    
    # calculate retention rate
    mutate(rate_transition=100*n_user/previous_segment_count)
    
  # labelling the transition
  user_retention_rate_data$transition<-paste(user_retention_rate_data$previous_segmentation,user_retention_rate_data$current_segmentation,sep='--')

  # export user retention result to csv
  write.csv(user_retention_rate_data,
            paste0("user_retention_summary_",
                   format(latest_date, '%Y'),
                   format(latest_date, '%m'),
                   format(latest_date, '%d'),
                   "_to_",
                   format(earliest_date+1, '%Y'),
                   format(earliest_date+1, '%m'),
                   format(earliest_date+1, '%d'),
                   ".csv"), row.names = FALSE)
  
  # return the data summary
  return(user_retention_rate_data)
  
}
```

```{r}
qiroah_user_segment_transition_summary<-retention_rate_summarization(qiroah_user_segment_summary, qiroah_user_retention)

head(qiroah_user_segment_transition_summary)

```

```{r}
# function to visualize the trend of a certain user retention 
user_retention_visualization<- function(user_retention_summary_data, selected_retention){
  
  # filter user segment data
  filtered_retention<-user_retention_summary_data %>% filter(transition==selected_retention)
  
  # identify range of data
  earliest_date<-format(min(filtered_retention$user_retention_period),'%Y-%m-%d')
  latest_date<-format(max(filtered_retention$user_retention_period),'%Y-%m-%d')
  
  # visualize a line chart
  filtered_retention %>%
    ggplot(aes(x=user_retention_period, y=rate_transition)) +
    geom_line(color="#69b3a2", size=1)+ 
    labs(title = "Qiroah App Learning Activity User Retention Trend",
         subtitle=paste0(selected_retention,
                         ', Period: ',earliest_date,' to ',latest_date),
         y='Rate',x='Date')
}
```

```{r}
# execute visualization in iteration of user retention segment
for(i in unique(qiroah_user_segment_transition_summary$transition)){
  print(user_retention_visualization(qiroah_user_segment_transition_summary,i))
}
```
